Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-12-06 survey/survey.tex
Remove material that is clearly not in the scope of the paper as currently outlined.
Move material which might still be useful to the end.

Some material which is more for us is clearly marked so.

\jc{wavefront}

Example : $2+5 = 5+2$. Green and Red apples versus apples. What
can we observe? If we can't observe something, then we call them
the same (Leibniz).

All programs, proofs, deductions are equivalences, isomorphisms,
reversible

Several pages on reversible logic; reversible family of level 1 and
level 2 languages (Pi);

We should also have a quick survey of other works in reversible
computation? Most of the other work tries to start from well known
models of computation or well known programming languages, and then
adapts them to be reversible. Pi is different in that it starts from
the semantic implications of reversibility. Then, because it adds
types, rather than control-flow (or even data-flow) as its next layer
of ``understanding'', this leads to equivalences, isomorphisms, etc.
This then leads, quite naturally, to finding that the ``proof
language'' of semirings (and Rig Groupoids at level 2) is actually a
programming language. And it is Pi. This is a neat twist on
Curry-Howard because CH is about \textbf{inhabitation} only. But with
``conservation of information'' as the basis, a different kind of
correspondance arises; in fact, this one may well be an actual
isomorphism.

\subsection{Equalities as Isomorphisms}

Translated to the computational world, the essence of our thesis is this:
\emph{to speak of equality is to speak of computation}. This view is partly
familiar from the study of various systems of equalities, called
\emph{calculi}, as models of computation. The \lcal, for instance, is a
system of equalities over a set of syntactic $\lambda$-terms such that they
result in a model of computation. We distill this idea to its most primitive
form. We show that there is computational content not just in certain
specific calculi, such as the \lcal, but even in our most primitive notion of
equality -- isomorphisms of finite sets.

The equalities we speak of are not \emph{a priori} extensional
notions, but are intensional specifications. To talk about two things
being equal, one has to show \emph{how they are equal} wherein this
\emph{how} is an intensional definition of a translation from
witnesses of the first to witnesses of the second and vice versa,
i.e. they are programs. Equivalences that do not have operational
content do not exist for us.

One may ask, what is the point of such a rarefied spartan view of
computation? How does this compare with more conventional models of
computation?  \emph{What is the point?}

\begin{enumerate}

\item The result is computation in its purest form: Programs written
  in this model are witnesses of equalities. In the process of
  computing these programs neither gain nor lose information. We get a
  crisp notion of quantitative information content of programs and
  show that \emph{computation preserves information}.

\item From the perspective of Quantum Physics, the physical world is
  one where every fundamental interaction is essentially reversible
  and various quantities such as energy, mass, angular momentum are
  conserved. Computation is no longer at odds with
  physics. Conservation of information and logical reversibility are
  intrinsic properties of computation as well. Further, various
  category theoretic models such as dagger symmetric traced monoidal
  categories studied for Quantum Physics are also the categorical
  models for computation based on isomorphisms.

\item Irreversibility is a computational effect: By embracing
  irreversible physical primitives, models of computation have also
  implicitly included a class of computational effects which we call
  \emph{information effects}. A consequence of our careful treatment
  of information, is that we effectively capture the gap between
  reversible computation and irreversible computation using a
  type-and-effect system.

What do we mean when we talk about intentional specification of
equality leading to an information preserving model of computation?
The equalities we talk of are processes to transform one thing to the
other. For example, if one were to believe that apples and oranges
were equal then one would have to show how any given apple can be
transformed to an orange i.e. witness of apple-hood get transformed to
witnesses of orange-hood (and vice versa). This is inherently
different from an extensional notion of equality such as
\emph{knowledge is power} and have a more operational sense. Equality
in this sense acts more like the the more general notion of
\emph{isomorphism} as probably best explained in the paper
``\emph{When is one thing equal to some other
  thing?}''~\cite{mazur2008one}.

In our formalism, we are not concerned with fruit but with very simple
things, namely numbers and sets. When we talk of a set of five apples,
we are not concerned about the individuality of the apples that
constitute the set. Rather, each apple in the set has the same status
as any other apple and we care about the individual apples only to the
extent that we can tell them apart to count five distinct ones. We
care about the abstract idea of fiveness and its equality with any
other abstract idea of fiveness.  When used as types, sets take the
role of possibility spaces -- we are no longer talking about a set of
five apples, but talking of one apple out of a possible five. This is
also where the switch from sets to information happens: When one apple
out of five is identified, it corresponds to the information to
discriminate between five choices. Clearly a choice of one out of ten
presents more information in the sense that it is a more
discriminating choice. When a computation promises an output of a
certain type, the resultant value is a witness of one choice out of
several. This is the sense in which we are concerned with equalities
of sets.

Programs in our model are built from descriptions of equalities or
sets/numbers. These are the familiar laws of arithmetic or logic:
\emph{Commutativity.} The generalized notion of swapping,
i.e. $a*b=b*a$ and $a+b=b+a$.
\emph{Associativity.} The
generalized notion of grouping, i.e. $a*(b*c)=(a*b)*c$ and
$a+(b+c)=(a+b)+c$.
\emph{Identity.} Multiplication by 1 and addition with 0,
  i.e. $a*1=a$ and $a+0=a$.
\emph{Distributivity.} Multiplication distributes over addition
  and cancels at zero, i.e. $a*(b+c)=a*b+a*c$ and $a*0=0$.
\emph{Cancellation.} Cancellation is the idea that equal things
  on both sides of a equality can be canceled out, i.e. $a+b=a+c$
  implies $b=c$, and is referred to as a \emph{trace} in category
  theory~\cite{joyal1996traced}.

How, one may ask, does this form a computational model? Even in very
simple computational models such as the SK combinatory basis, the
combinator K provides the basis of choice/conditional by deleting an
argument and the combinator S provides the basis for iteration by
duplicating an argument. As we will see, distributivity gives us
conditionals and trace gives us iteration, albeit in a logically
reversible manner.  In Chs.~\ref{ch:pi} and~\ref{ch:pi0} we work out
the details of the computational model, show that it is Turing
complete and develop programs in it including several simple numeric
operations, algebraic manipulations of trees and a meta-circular
interpreter (see Ch.~\ref{ch:interpreters}).

More formally, we develop an information preserving model of
computation, wherein the process of computing does not gain or lose
information. Our model arises from a computational interpretation of
type isomorphisms with iso-recursive types and trace. This can
equivalently be described as computing a commutative semi-ring with
cancellation or as a traced dagger symmetric bimonoidal category. The
category has a groupoid structure and all computations are logically
reversible.

To summarize, the most primitive notion of equality that is available
to us -- that of isomorphisms -- already contains the notion of
computation. The resulting computational model exposes the fine
structure of how computation handles information. We believe that such
a model of computing is foundational and we are only beginning to
understand its applications and full potential. In conclusion, let us
quote Paul Blain Levy's fine advocacy slogan: Once the fine structure
has been exposed, why ignore it?

%%%
\subsection{Information}

We can now describe information preservation in a precise sense.  Let
`$b$' be a (not necessarily finite) type whose values are labeled
$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
indicates the probability of the output of the function to have value
${b_2}^j$. We say a function is \emph{information-preserving} if its
output entropy is equal to the entropy of its input. See
Ch.~\ref{ch-metalang} for details.

Operations that change the information content of programs have a
special status. We call them \emph{information effects} and they are
encapsulated using an arrow meta-language in much the same way that
one would encapsulate other computational effects in the \lcal\ using a
monadic metalanguage.  The treatment of information effects is
analogous to open and closed systems in physics. Closed physical
systems conserve mass and energy and are the basic unit of study in
physics. Pure computations that do not have information effects are
like closed physical systems. They describe equalities.  Open systems
interact with their environment, possibly exchanging mass or
energy. These interactions may be thought of as \emph{effects} that
modify the conservation properties of the system. Computations with
information effects are much like open systems and they can be
converted into pure computations by making explicit the surrounding
information environment that they interact with.

Three things follow: (1) We capture the gap between reversible and
irreversible computation with a type-and-effect system. (2)
Categorically speaking, information effects turn monoidal categories
into ones that have a cartesian structure corresponding to
conventional computation.  (3) We show how conventional irreversible
computation such as the \lcal\ can be embedded into this model, such
that the embedding makes the implicit information effects of \lcal\
explicit.  As a consequence of this approach, many applications in
which information manipulation is computationally significant are put
within the reach of our conceptual model of computation. Such
applications include quantitative information-flow
security~\cite{myerssab}, differential
privacy~\cite{dwork:differential}, energy-aware
computing~\cite{1324180,605411}, VLSI
design~\cite{Macii:1996:ECE:874066.875828}, and biochemical models of
computation~\cite{bio}.

Ch.~\ref{ch:piee} focuses on work in progress. It deals with the
duality of computation in the setting of information
preservation. Andrzej Filinski's work on the symmetric
\lcal\ established the fascinating connection that values are dual to
continuations, functions are dual to delimited continuations and
established the duality of call-by-name and call-by-value evaluation
strategies~\cite{Filinski:1989:DCI:648332.755574}. We explore
Filinski-style duality in the setting of our information preserving
model of computing. Here duality does not appear as one De
Morgan-style dualizing $\neg$ operator, but as two distinct dualizing
operations -- negation and division. This results in a crisp semantics
for negative and fractional types and the resulting system has a
equational formalism in zero-totalized fields or
meadows~\cite{DBLP:journals/tcs/BergstraHT09}. Computational analogues
of physical phenomena such as superposition and entanglement appear in
the resulting semantics.

Talk about conservation of information in biology, computer science,
evolutionary computing, and search. (See
\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.

%%%
\subsection{Technical Overview}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programs as Reversible Deformations}

@@

>>>>

\amr{We should also have a quick survey of other works in reversible
computation? Most of the other work tries to start from well known
models of computation or well known programming languages, and then
adapts them to be reversible. Pi is different in that it starts from
the semantic implications of reversibility. Then, because it adds
types, rather than control-flow (or even data-flow) as its next layer
of ``understanding'', this leads to equivalences, isomorphisms, etc.
This then leads, quite naturally, to finding that the ``proof
language'' of semirings (and Rig Groupoids at level 2) is actually a
programming language. And it is Pi. This is a neat twist on
Curry-Howard because CH is about \textbf{inhabitation} only. But with
``conservation of information'' as the basis, a different kind of
correspondance arises; in fact, this one may well be an actual
isomorphism.}

\subsection{Information}

We can now give a first precise definition of information preservation.
Let
`$b$' be a (not necessarily finite) type whose values are labeled
$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
indicates the probability of the output of the function to have value
${b_2}^j$. We say a function is \emph{information-preserving} if its
output entropy is equal to the entropy of its input. In later sections,
we will refine this view.

\amr{
Talk about conservation of information in biology, computer science,
evolutionary computing, and search. (See
\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
}

%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programs as Reversible Deformations}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Below are pieces of text that are not immediately useful above, but
%% from which we might still be able to pilfer usefully.
\subsection{Technical Overview}
